{\sc Kalashnikov}: Fully Automatic Loop Free Program Synthesis by Bounded Model Checking

### 2013-10-02 paper/paper.tex
Added 2sum algorithm & 2sum timings
Added more about C-, schematic diagrams

@@

>>>>

\makeatletter
\pgfdeclareshape{datastore}{
  \inheritsavedanchors[from=rectangle]
  \inheritanchorborder[from=rectangle]
  \inheritanchor[from=rectangle]{center}
  \inheritanchor[from=rectangle]{base}
  \inheritanchor[from=rectangle]{north}
  \inheritanchor[from=rectangle]{north east}
  \inheritanchor[from=rectangle]{east}
  \inheritanchor[from=rectangle]{south east}
  \inheritanchor[from=rectangle]{south}
  \inheritanchor[from=rectangle]{south west}
  \inheritanchor[from=rectangle]{west}
  \inheritanchor[from=rectangle]{north west}
  \backgroundpath{
    %  store lower right in xa/ya and upper right in xb/yb
    \southwest \pgf@xa=\pgf@x \pgf@ya=\pgf@y
    \northeast \pgf@xb=\pgf@x \pgf@yb=\pgf@y
    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@ya}}
    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@ya}}
    \pgfpathmoveto{\pgfpoint{\pgf@xa}{\pgf@yb}}
    \pgfpathlineto{\pgfpoint{\pgf@xb}{\pgf@yb}}
 }
}
\makeatother

@@

\begin{figure}
\begin{center}
\begin{tikzpicture}[
  scale=0.8,
  font=\sffamily,
  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
  process/.style={draw,thick,circle,fill=blue!20},
  sink/.style={source,fill=green!20},
  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
  dots/.style={gray,scale=2},
  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
  every node/.style={align=center, scale=0.8}]

  % Position the nodes using a matrix layout
  \matrix{
    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
    \& \node[datastore] (tests) {\texttt{tests.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
    \node[datastore] (cbmc-synth) {\texttt{cbmc/synth.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-synth) {\texttt{explicit/synth.c}}; \\
    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
    \& \node[sink] (cbmc-out) {candidate program}; \& \node[sink] (explicit-out) {candidate program}; \& \\
  };

  \draw[to] (driver) -- (tests);
  \draw[to] (user) -- (spec);

  \draw[to] (tests) -- (cbmc);
  \draw[to] (spec) -- (cbmc);
  \draw[to] (interpreter) -- (cbmc);

  \draw[to] (tests) -- (gcc);
  \draw[to] (spec) -- (gcc);
  \draw[to] (interpreter) -- (gcc);

  \draw[to] (cbmc-synth) -- (cbmc);
  \draw[to] (explicit-synth) -- (gcc);

  \draw[to] (gcc) -- (aout);

  \draw[to] (cbmc) -- (cbmc-out);
  \draw[to] (aout) -- (explicit-out);
\end{tikzpicture}
\end{center}

\label{fig:synth-dfd}
\caption{Schematic diagram of {\sc synth}}
\end{figure}

\begin{figure}
\begin{center}
\begin{tikzpicture}[
  scale=0.8,
  font=\sffamily,
  every matrix/.style={ampersand replacement=\&,column sep=.5cm,row sep=.75cm},
  source/.style={draw,thick,rounded corners,fill=yellow!20,inner sep=.3cm},
  process/.style={draw,thick,circle,fill=blue!20},
  sink/.style={source,fill=green!20},
  datastore/.style={draw,very thick,shape=datastore,inner sep=.3cm},
  dots/.style={gray,scale=2},
  to/.style={->,>=stealth',shorten >=1pt,semithick,font=\sffamily\footnotesize},
  every node/.style={align=center, scale=0.8}]

  % Position the nodes using a matrix layout
  \matrix{
    \& \node[source] (driver) {driver}; \& \node[source] (user) {user}; \& \\
    \& \node[datastore] (prog) {\texttt{prog.c}}; \& \node[datastore] (spec) {\texttt{spec.c}}; \& \node[datastore] (interpreter) {\texttt{interpreter.c}}; \\
    \node[datastore] (cbmc-verif) {\texttt{cbmc/verif.c}}; \& \node[process] (cbmc) {\sc cbmc}; \& \node[process] (gcc) {\sc gcc}; \& \node[datastore] (explicit-verif) {\texttt{explicit/verif.c}}; \\
    \& \& \node[process] (aout) {\texttt{a.out}}; \& \\
    \& \node[sink] (cbmc-out) {test vector}; \& \node[sink] (explicit-out) {test vector}; \& \\
  };

  \draw[to] (driver) -- (prog);
  \draw[to] (user) -- (spec);

  \draw[to] (prog) -- (cbmc);
  \draw[to] (spec) -- (cbmc);
  \draw[to] (interpreter) -- (cbmc);

  \draw[to] (tests) -- (gcc);
  \draw[to] (spec) -- (gcc);
  \draw[to] (interpreter) -- (gcc);

  \draw[to] (cbmc-verif) -- (cbmc);
  \draw[to] (explicit-verif) -- (gcc);

  \draw[to] (gcc) -- (aout);

  \draw[to] (cbmc) -- (cbmc-out);
  \draw[to] (aout) -- (explicit-out);
\end{tikzpicture}
\end{center}

\label{fig:synth-dfd}
\caption{Schematic diagram of {\sc verif}}
\end{figure}

\section{C-}
We use a fragment of C as our underlying logic.  We will refer to this fragment
as C-.  The characteristic property of C- is that safety can be decided for
C- programs by using a bounded model checker.  A C- program is just a C program
with the following syntactic restrictions:

\begin{itemize}
 \item All loops in the program must have a constant bound.
 \item All recursion in the program must be limited to a constant depth.
 \item All arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
\end{itemize}

Additionally, C- programs may use:

\begin{itemize}
 \item Nondeterministic values.
 \item Assumptions.
 \item Arbitrary width types.
\end{itemize}

So for example the programs in Fig.~\ref{fig:c-} are valid
C- programs, but the program in Fig.~\ref{fig:not-c-} is not, since it uses
a non-constant loop bound.

\begin{figure}
\begin{minipage}[scale=0.8]{0.45\linewidth}
 \begin{lstlisting}[language=c]
int count_bits(int x) {
  int i, ret;

  ret = 0;

  for (i = 0; i < 32; i++) {
    if (x & (1 << i))
      ret++;
  }

  return ret;
}
 \end{lstlisting}
\end{minipage}
\begin{minipage}{0.54\linewidth}
 \begin{lstlisting}[language=C]
int common_factor(int A[10]) {
  int factor = nondet();
  int i;

  for (i = 0; i < 10; i++) {
    assume((A[i] % factor) == 0);
  }

  assume(factor > 1);
  return factor;
}

 \end{lstlisting}
\end{minipage}

 \label{fig:c-}
 \caption{Two C- programs.}
\end{figure}

\begin{figure}
 \begin{lstlisting}[language=C]
int fib(int n) {
  int a, b, c;

  a = b = 1;

  while (n--) {
    c = a+b;
    a = b;
    b = c;
  }

  return a;
}
 \end{lstlisting}

 \label{fig:not-c-}
 \caption{A C program to compute Fibonacci numbers.}
\end{figure}

Since each loop is bounded by a constant, and each recursive function call is
limited to a cosntant depth, a C- program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound $k$, then
a bounded model checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a C- program of
size $l$ and with largest loop bound $k$, a bounded model checker will
create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop free C- program
of size $O(s)$.  The safety problem for C- is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.

\subsection{Why use C- instead of SAT?}
To misquote David Wheeler: ``all problems in computer science can be solved by
another level of abstraction.''~\cite{wheeler-abstraction}.  C- serves as
an abstraction of SAT.  We can think of SAT as an assembly language and
{\sc cbmc} as a compiler for C- to this assembly language.  The benefits
of using C- as our logic rather than SAT are very similar to the benefits of
writing code in a high level language rather than assembly language.
Some of the more significant advantages are:

\begin{itemize}
 \item The code is more readable and easier to debug.
 \item The code is much more concise.
 \item We can take advantage of new optimisations, different model
 checkers and different backends (SAT, SMT, ACDCL, etc.) for free.
\end{itemize}

To illustrate some of these points, we were able to implement a first
version of {\sc kalashnikov} in around 150 lines of code, which took
around two hours to write.  The resulting code was easy to modify,
optimise and debug.  The current version, which includes floating point
support, several optimisations and two model checking algorithms,
is still less than 1200 lines of code (600 lines of C- and 600 lines
of Python).

Another benefit of using C- as our logic is that it would be easy to
automatically extract specifications from C programs, which would be
useful in the context of superoptimisation.  This is because C- is
a relatively large fragment of C, so we anticipate that many interesting
sections of real code will already be C-.
